---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# marx

<!-- badges: start -->

<!-- badges: end -->

`marx` is an R package that implements [Ma]{.underline}trix factorization and [R]{.underline}esidual E[x]{.underline}pression (MARX), an algorithm for comparing high-dimensional biological measurements to a lower-dimensional reference linear subspace. Currently, MARX is unpublished and highly experimental.

## Installation

You can install the development version of marx from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("keyes-timothy/marx")
```

## Example

This is a basic example for using `marx` to decompose an input dataset into two components: one that aligns with a reference dataset (represented as a linear subspace) and one that doesn't align with the reference.

```{r example}
library(marx)

# simulate reference data 
# this data is assumed to be an expression matrix in which rows represent 
# observations and columns represent measurements (i.e. genes). 
healthy_data <-
  matrix(data = runif(n = 300 * 40), nrow = 300, ncol = 40) |>
  dplyr::as_tibble()

# simulate new data 
disease_data <- 
  matrix(data = runif(n = 700 * 40, min = 1, max = 2), nrow = 700, ncol = 40) |>
  dplyr::as_tibble()

# build a healthy reference subspace using the SVD
healthy_subspace <- 
  healthy_data |> 
  # other method options include "nmf" and "pma"
  marx_find_healthy_subspace(pca_threshold = 0.9, method = "svd")

# project the disease_data onto the healthy subspace
disease_projection <- 
  disease_data |> 
  marx_project(healthy_subspace = healthy_subspace)

head(disease_projection)
```

marx can also compute the lineage- and disease-specific component magnitudes (i.e. the reconstruction and the reconstruction loss for each observation using the reference subspace):

```{r}
disease_projection |> 
  marx_find_magnitudes() |> 
  head()
```

We can compare the disease-specific component of the healthy data and the disease data:

```{r}
# find healthy projection
healthy_projection <- 
  healthy_data |> 
  marx_project(healthy_subspace = healthy_subspace)

# find healthy magnitudes 
healthy_magnitudes <- 
  healthy_projection |> 
  marx_find_magnitudes()

disease_magnitudes <- 
  disease_projection |> 
  marx_find_magnitudes()

plot_tibble <- 
  dplyr::bind_rows(
    dplyr::mutate(healthy_magnitudes, sample_type = "healthy"), 
    dplyr::mutate(disease_magnitudes, sample_type = "disease")
  )

plot_tibble |> 
  ggplot2::ggplot(ggplot2::aes(x = disease_magnitude, fill = sample_type)) + 
  ggplot2::geom_density(alpha = 0.5) + 
  ggplot2::theme_bw() + 
  ggplot2::labs(subtitle = "diseased tissue has a larger disease-specific component than healthy tissue")
```
